---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# epitree

<!-- badges: start -->
[![R-CMD-check](https://github.com/CyGei/epitree/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/CyGei/epitree/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `epitree` package provides a statistical framework for comparing posterior distributions of transmission trees (*and soon*, phylogenetic trees) generated through Bayesian inference. It employs PERMANOVA (via [`vegan::adonis2`](https://vegandevs.github.io/vegan/reference/adonis.html)) to assess whether trees from different models or MCMC chains stem from the same underlying distribution.

## Installation

You can install the development version of epitree from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("CyGei/epitree")
```

# Quick start

```{r}
library(epitree)

# Simulate two chains of transmission trees
chainA <- lapply(1:100, function(i) {
  make_tree(20, R = 2, stochastic = TRUE) |>
    igraph::as_long_data_frame()
})
chainB <- lapply(1:100, function(i) {
  make_tree(20, R = 4, stochastic = TRUE) |>
    igraph::as_long_data_frame()
})

# Compare the two chains
result <- compare_chains(chainA, chainB)
print(result)

```

